---
title: "Quantifying Clonal Bias"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{Quantifying Clonal Bias}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
p.caption {
  font-size: 0.9em;
}
</style>
 
```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)

suppressMessages(library(scRepertoire))
suppressMessages(library(Seurat))
data("contig_list") 
combined.TCR <- combineTCR(contig_list, 
                           samples = c("P17B", "P17L", "P18B", "P18L", 
                                            "P19B","P19L", "P20B", "P20L"))

scRep_example <- readRDS("scRep_example_full.rds")

scRep_example <- combineExpression(combined.TCR, 
                                   scRep_example, 
                                   cloneCall="gene", 
                                   group.by = "sample")

#Adding patient information
scRep_example$Patient <- substr(scRep_example$orig.ident, 1,3)

#Adding type information
scRep_example$Type <- substr(scRep_example$orig.ident, 4,4)

#Defining colors
colorblind_vector <- hcl.colors(n=7, palette = "inferno", fixup = TRUE)
```


## StartracDiversity

From the excellent work by [Lei Zhang, et al.](https://www.nature.com/articles/s41586-018-0694-x), the authors introduced new methods for looking at clones by cellular origins and cluster identification. Their [STARTRAC](https://github.com/Japrin/STARTRAC) software has been adapted to work with `scRepertoire` and please read and cite their excellent work. 

To use the `StartracDiversity()` function, you need to include the product of the `combineExpression()` function. The second requirement is a column header in the metadata of the Seurat object that has the tissue of origin. In the example data, type corresponds to the column "Type", which includes the "P" (peripheral blood) and "L" (lung) classifiers. The indices can be subsetted for a specific patient or examined overall using the group.by variable. Importantly, the function uses only the strict definition of a clone, which includes the VDJC genes and the CDR3 nucleotide sequence.

### Indices Output from `StartracDiversity()`

*  `expa` - Clonal Expansion  
*  `migr` - Cross-tissue Migration  
*  `tran` - State Transition  

Key Parameters for `StartracDiversity()`

* `type`: The variable in the metadata that provides tissue type.
* `group.by`: A column header in the metadata to group the analysis by (e.g., "sample", "treatment"). 

Calculate and plot Startrac diversity metrics using "Type" for tissue and "Patient" for grouping:

```{r tidy = FALSE}
StartracDiversity(scRep_example, 
                  type = "Type", 
                  group.by = "Patient")
```

## clonalBias

A new metric proposed by [Massimo et al](https://pubmed.ncbi.nlm.nih.gov/35829695/), ```clonalBias()```, like STARTRAC, is a clonal metric that seeks to quantify how individual clones are skewed towards a specific cellular compartment or cluster. A clone bias of `1` indicates that a clone is composed of cells from a single compartment or cluster, while a clone bias of `0` matches the background subtype distribution. Please read and cite the linked manuscript if using `clonalBias()`

Key Parameter(s) for `clonalBias()`

* `group.by`: A column header in the metadata that bias will be based on.
* `split.by`: The variable to use for calculating the baseline frequencies (e.g., "Type" for lung vs peripheral blood comparison)
* `n.boots`: Number of bootstraps to downsample.
* `min.expand`: Clone frequency cut-off for the purpose of comparison (default = 10).

Here we calculate and plot clonal bias using `aa` clone calls, splitting by "Patient" and grouping by "seurat_clusters", with a minimum expansion of 5 and 10 bootstraps:

```{r, message = FALSE, tidy = FALSE}
clonalBias(scRep_example, 
           cloneCall = "aa", 
           split.by = "Patient", 
           group.by = "seurat_clusters",
           n.boots = 10, 
           min.expand =5)
```
