---
title: "Basic Clonal Visualizations"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{Basic Clonal Visualizations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
p.caption {
  font-size: 0.9em;
}
</style>
 
```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)

suppressMessages(library(scRepertoire))
data("contig_list") 
combined.TCR <- combineTCR(contig_list, 
                           samples = c("P17B", "P17L", "P18B", "P18L", 
                                            "P19B","P19L", "P20B", "P20L"))
combined.TCR <- addVariable(combined.TCR, 
                            variable.name = "Type", 
                            variables = rep(c("B", "L"), 4))
```

## clonalQuant

The first function to explore the clones is `clonalQuant()` to return the total or relative numbers of unique clones. 

**scale**  

* TRUE - relative percent of unique clones scaled by the total size of the clonal repertoire.
* FALSE - Report the total number of unique clones **(default)**.

**chain**  

+ "both" for combined chain visualization  
+ "TRA", "TRB", "TRD", "TRG", "IGH" or "IGL" to select single chain

```{r tidy = FALSE}
clonalQuant(combined.TCR, 
            cloneCall="strict", 
            chain = "both", 
            scale = TRUE)
```

Another option here is to be able to define the visualization by data classes. Here, we used the `combineTCR()` list to define the **Type** variable as part of the naming structure. We can use the **group.by** to specifically use a column in the data set to organize the visualization.

```{r}
clonalQuant(combined.TCR, 
            cloneCall="gene", 
            group.by = "Type", 
            scale = TRUE)
```

## clonalAbundance

We can also examine the relative distribution of clones by abundance. Here `clonalAbundance()` will produce a line graph with a total number of clones by the number of instances within the sample or run. Like above, we can also group.by this by vectors within the contig object using the **group.by** variable in the function.

```{r tidy = FALSE}
clonalAbundance(combined.TCR, 
                cloneCall = "gene", 
                scale = FALSE)
```

`clonalAbundance()` output can also be converted into a density plot, which may allow for better comparisons between different repertoire sizes, by setting **scale** = TRUE. 
```{r}
clonalAbundance(combined.TCR, 
                cloneCall = "gene", 
                scale = TRUE)
```


## clonalLength

We can look at the length distribution of the CDR3 sequences by calling the `lengtheContig()` function. Importantly, unlike the other basic visualizations, the **cloneCall** can only be **"nt"** or **"aa"**. Due to the method of calling clones as outlined above, the length should reveal a multimodal curve, this is a product of using the **NA** for the unreturned chain sequence and multiple chains within a single barcode. 

**chain**  

+ "both" for combined chain visualization  
+ "TRA", "TRB", "TRD", "TRG", "IGH" or "IGL" to select single chain  

```{r tidy = FALSE}
clonalLength(combined.TCR, 
             cloneCall="aa", 
             chain = "both") 

clonalLength(combined.TCR, 
             cloneCall="aa", 
             chain = "TRA", 
             scale = TRUE) 
```

## clonalCompare

We can also look at clones between samples and changes in dynamics by using the `clonalCompare()` function.

**samples**  

* Can be used to isolate specific samples based on the name of the list element

**graph**  

* "alluvial" - graph imaged below  
* "area" - graph by area of the respective clone  

**top.clones**  

* The top number of clones to graph, this will be calculated based on the frequency of the individual sample. This can also be left blank.  

**clones**  

* Can be used to isolate specific clone sequences, ensure the call matches the sequences you would like to visualize.  

**highlight.clones**  

* Specifically color certain clones, other clones will be greyed out.

**relabel.clones**  

* Simplify the isolated clones to numerical designations to allow for a tidier visualization

```{r tidy = FALSE}
clonalCompare(combined.TCR, 
                  top.clones = 10, 
                  samples = c("P17B", "P17L"), 
                  cloneCall="aa", 
                  graph = "alluvial")
```

We can also choose to highlight specific clones, such as in the case of *"CVVSDNTGGFKTIF_CASSVRRERANTGELFF"* and *"NA_CASSVRRERANTGELFF"* using the **highlight.clones** parameter. In addition, we can simplify the plot to label the clones as clones 1:19.

```{r tidy = FALSE}
clonalCompare(combined.TCR, 
              top.clones = 10,
              highlight.clones = c("CVVSDNTGGFKTIF_CASSVRRERANTGELFF", "NA_CASSVRRERANTGELFF"),
              relabel.clones = TRUE,
              samples = c("P17B", "P17L"), 
              cloneCall="aa", 
              graph = "alluvial")
```

Alternatively, if we only want to show specific clones, we can use the **clones** parameter.

```{r}
clonalCompare(combined.TCR, 
              clones = c("CVVSDNTGGFKTIF_CASSVRRERANTGELFF", "NA_CASSVRRERANTGELFF"),
              relabel.clones = TRUE,
              samples = c("P17B", "P17L"), 
              cloneCall="aa", 
              graph = "alluvial")
```


## clonalScatter

```clonalScatter()``` will organize two repertoires, quantify the relative clone sizes, and produce a scatter plot comparing the two samples. 

**x.axis** and **y.axis** 

* Names of the list element to place on the x-axis and y-axis - so for example *"P18B"* and *"P18L"* 

**dot.size**  

* "total" to display the total number of clones between the x- and y-axis  
* Name of the list element to use for size calculation  
 
**graph**  

* "proportion" for the relative proportion the clone represents across all clones  
* "count" for the total count of clones by sample  

```{r tidy = FALSE}
clonalScatter(combined.TCR, 
              cloneCall ="gene", 
              x.axis = "P18B", 
              y.axis = "P18L",
              dot.size = "total",
              graph = "proportion")
```