---
title: "Quantifying Clonal Bias"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{Quantifying Clonal Bias}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
p.caption {
  font-size: 0.9em;
}
</style>
 
```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)

suppressMessages(library(scRepertoire))
suppressMessages(library(Seurat))
data("contig_list") 
combined.TCR <- combineTCR(contig_list, 
                           samples = c("P17B", "P17L", "P18B", "P18L", 
                                            "P19B","P19L", "P20B", "P20L"))

scRep_example <- readRDS("scRep_example_full.rds")

scRep_example <- combineExpression(combined.TCR, 
                                   scRep_example, 
                                   cloneCall="gene", 
                                   group.by = "sample")

#Adding patient information
scRep_example$Patient <- substr(scRep_example$orig.ident, 1,3)

#Adding type information
scRep_example$Type <- substr(scRep_example$orig.ident, 4,4)

#Defining colors
colorblind_vector <- hcl.colors(n=7, palette = "inferno", fixup = TRUE)
```


## StartracDiversity

From the excellent work by [Lei Zhang, et al.](https://www.nature.com/articles/s41586-018-0694-x), the authors introduce new methods for looking at clones by cellular origins and cluster identification. Their [STARTRAC](https://github.com/Japrin/STARTRAC) software has been adapted to work with scRepertoire and please read and cite their excellent work. 

In order to use the ```StartracDiversity()``` function, you will need to include the product of the ```combinedExpression()``` function. The second requirement is a column header in the meta data of the Seurat object that has tissue of origin. In the example data, **type** corresponds to the column "Type", which includes the "P" and "T" classifiers. The indices can be subsetted for a specific patient or examined overall using the **by** variable. Importantly, the function uses only the strict definition of a clone of the VDJC genes and the CDR3 nucleotide sequence. 

**The indices output includes:**  

*  expa - Clonal Expansion  
*  migr - Cross-tissue Migration  
*  tran - State Transition  

```{r tidy = FALSE}
StartracDiversity(scRep_example, 
                  type = "Type", 
                  group.by = "Patient")
```

## clonalBias

A new metric proposed by [Massimo et al](https://pubmed.ncbi.nlm.nih.gov/35829695/), ```clonalBias()```, like STARTRAC is a clonal metric that seeks to quantify how individual clones are skewed towards a specific cellular compartment or cluster.

**split.by**   

* Variable used to calculate the baseline frequencies  

**group.by**  

* The compartment/variable used for the purpose of the comparison  

**min.expand**  

* Cut point for frequency **(default = 10)**

```{r, message = FALSE, tidy = FALSE}
clonalBias(scRep_example, 
           cloneCall = "aa", 
           split.by = "Patient", 
           group.by = "seurat_clusters",
           n.boots = 10, 
           min.expand =5)
```
