---
title: "Comparing Clonal Diversity and Overlap"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{Comparing Clonal Diversity and Overlap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
***

<style>
p.caption {
  font-size: 0.9em;
}
</style>
 
```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)

suppressMessages(library(scRepertoire))
data("contig_list") 
combined.TCR <- combineTCR(contig_list, 
                           samples = c("P17B", "P17L", "P18B", "P18L", 
                                            "P19B","P19L", "P20B", "P20L"))
```

## clonalDiversity

Diversity can also be measured for samples or by other variables. Diversity **metrics** calculated, include: **"shannon"**, **"inv.simpson"**, **"norm.entropy"**, **"gini.simpson"**, **"chao1"**, and **"ACE"**. Please see the manual for more information on each metric and the underlying calculations. 

Inherent in diversity calculations is a bias increasing diversity with increasing repertoire size. ```clonalDiversity()``` will automatically downsample to the smallest repertoire size and perform boostapping to return the mean diversity estimates. If the output of diversity values are strange or minimally variable, it is likely due to a sample with small repertoire size.

**n.boots**  
The number of calculations to perform **(default = 100)**.  

**return.boots**  
- TRUE: Return all the calculations.  
- FALSE: Return only the mean values **(default)**.  

**skip.boots**  
Skip the bootstrapping calculations.


```{r}
clonalDiversity(combined.TCR, 
                cloneCall = "gene", 
                n.boots = 20)

#Return only a subset of metrics
clonalDiversity(combined.TCR, 
                metrics = c("shannon", "ACE"),
                cloneCall = "gene", 
                n.boots = 20)
```
## clonaRarefaction

We can also use Hill numbers to estimate the rarefaction, or estimating species richness, using the the abundance of clones across groupings. Underlying the rarefaction calculation is the use of observed receptor of abundance to compute diversity. 

**hill.numbers** 
+ 0 - species-richness  
+ 1 - Shannon  
+ 2 - Simpson  

**plot.type**  
+ 1 - sample-size-based rarefaction/extrapolation  
+ 2 - sample completeness curve  
+ 3 - coverage-based rarefaction/extrapolation curve  

This relies on the [iNEXT](https://cran.r-project.org/web/packages/iNEXT/index.html) with the accompanying [manuscript](http://chao.stat.nthu.edu.tw/wordpress/paper/120_pdf_appendix.pdf). Like the other wrapping functions in scRepetoire, please cite the original work. 

### Rarefaction using Shannon Diversity (q = 0)

```{r, message=FALSE}
clonalRarefaction(combined.TCR,
                  plot.type = 1,
                  hill.numbers = 0,
                  n.boots = 2)

clonalRarefaction(combined.TCR,
                  plot.type = 2,
                  hill.numbers = 0,
                  n.boots = 2)

clonalRarefaction(combined.TCR,
                  plot.type = 3,
                  hill.numbers = 0,
                  n.boots = 2)
```

### Rarefaction using Shannon Diversity (q = 1)

```{r}
clonalRarefaction(combined.TCR,
                  plot.type = 1,
                  hill.numbers = 1,
                  n.boots = 2)

clonalRarefaction(combined.TCR,
                  plot.type = 2,
                  hill.numbers = 1,
                  n.boots = 2)

clonalRarefaction(combined.TCR,
                  plot.type = 3,
                  hill.numbers = 1,
                  n.boots = 2)
```

## clonalSizeDistribution

Another method for modeling the repertoire distribution is a discrete gamma-GPD spliced threshold model, proposed by [Koch et al.](https://pubmed.ncbi.nlm.nih.gov/30485278/) The spliced model models the repertoire and allows for the application of a power law distribution for larger clonal-expanded sequences and a Poisson distribution for smaller clones. After fitting the models, repertoires can be compared using Euclidean distance. 

If using this function, please read/cite [Koch et al.](https://pubmed.ncbi.nlm.nih.gov/30485278/) and check out the [powerTCR](https://bioconductor.org/packages/release/bioc/html/powerTCR.html) R package.

```{r}
clonalSizeDistribution(combined.TCR, 
                       cloneCall = "aa", 
                       method= "ward.D2")
```

## clonalOverlap

If you are interested in measures of similarity between the samples loaded into scRepertoire, using `clonalOverlap()` can assist in the visualization. 

```{r}
clonalOverlap(combined.TCR, 
              cloneCall = "strict", 
              method = "morisita")

clonalOverlap(combined.TCR, 
              cloneCall = "strict", 
              method = "raw")
```