---
title: "Frequently Asked Questions"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{Frequently Asked Questions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
p.caption {
  font-size: 0.9em;
}
</style>
 
```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = FALSE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)

suppressMessages(library(scRepertoire))
data("contig_list") 
combined.TCR <- combineTCR(contig_list, 
                           samples = c("P17B", "P17L", "P18B", "P18L", 
                                            "P19B","P19L", "P20B", "P20L"))
```

## Barcode Mismatch

Depending on the pipeline used to generate the single-cell object, there may be inherent mismatches in the barcodes in the single-cell object and the output of ```combineBCR()``` or ```combineTCR()```. 

### Common Barcode Modification Examples

**Seurat-modified barcodes**: By default, Seurat will append _X to the suffix of the barcodes.

* `original: ACGTACGTACGTACGT-1`
* `seurat-modified: ACGTACGTACGTACGT-1_1`

**scRepertoire-modified barcodes**: `scRepertoire` uses the samples and/or ID parameters in `combineTCR()` or `combineBCR()` to add a prefix to the barcodes.

* `original: ACGTACGTACGTACGT-1`
* `scRepertoire-modified: Sample1_ACGTACGTACGTACGT-1`
    
### Solution: Renaming Cell Barcodes in Seurat

The easiest way to make these compatible is to rename the cell barcodes in the Seurat object using `RenameCells()` from the `SeuratObject` package

```{r, eval=FALSE}
# Assuming 'seuratObj' is your Seurat object
cell.barcodes <- rownames(seuratObj[[]])
# removing the _1 at the end of the barcodes (adjust regex if your suffix differs)
cell.barcodes <- stringr::str_split(cell.barcodes, "_", simplify = TRUE)[,1]
# adding the prefix of the orig.ident to the barcodes, assuming that is the sample IDs
cell.barcodes <- paste0(seuratObj$orig.ident, "_", cell.barcodes)
seuratObj <- RenameCells(seuratObj, new.names = cell.barcodes)
```

## Adjusting Color Palettes

For all visualizations in `scRepertoire`, you have two primary ways to adjust the color scheme.

Methods for Color Adjustment

* **Internal Palette Selection**: Change the `palette` parameter within `scRepertoire` functions to the desired color scheme. This approach uses the built-in palettes of `grDevices`, and you can access the list of available color schemes using `hcl.pals()`. 
* **Adding a ggplot Layer**: Extend the `scRepertoire` plot (which is a ggplot object) by adding a `ggplot2` layer with a new color scheme using `scale_fill_manual()` or similar functions.  

### Using Internal Palette Selection

```{r}
# Internal Palette Selection
clonalQuant(combined.TCR, 
            cloneCall="strict", 
            chain = "both", 
            scale = TRUE, 
            palette = "Zissou 1")
```

### Using ggplot2 System:

```{r}
# Using gg System
clonalQuant(combined.TCR, 
            cloneCall="strict", 
            chain = "both", 
            scale = TRUE) + 
  scale_fill_manual(values = hcl.colors(8,"geyser"))
```

### Adjusting Plot Theme

Since `scRepertoire` functions return `ggplot` objects, modifying the general appearance or theme of the plot is straightforward, similar to adjusting color palettes—by adding a `ggplot2` theme layer.

```{r}
# Original clonalQuant plot
clonalQuant(combined.TCR, 
            cloneCall="strict", 
            chain = "both", 
            scale = TRUE)

# Modifying the theme of the clonalQuant plot
clonalQuant(combined.TCR, 
            cloneCall="strict", 
            chain = "both", 
            scale = TRUE) + 
  theme_classic()
```

### Adjusting Order of Plotting

The order of grouping/group.by variables in `scRepertoire` plots (whether along an axis or in color legends) can be precisely controlled using the order.by parameter.

Key Parameter for Plot Order

* `order.by`: A character vector defining the desired order of elements for the `group.by` variable. It's crucial that the strings in this vector exactly match the `group.by` strings. Alternatively, setting `order.by = "alphanumeric"` will automatically sort groups alphanumerically.

```{r}
clonalQuant(combined.TCR, 
            cloneCall="strict", 
            chain = "both", 
            scale = TRUE, 
            order.by = c("P17B","P18B","P19B","P20B","P17L","P18L","P19L","P20L"))
```


### Getting Data Used in Plots

Within each of the general analysis functions in `scRepertoire`, there's an option to export the underlying data frame used to create the visualization.

Key Parameter for Data Export

* `exportTable`: Set this parameter to `TRUE` to return the data frame used to generate the graph instead of the visual output.

```{r}
clonalQuant_output <- clonalQuant(combined.TCR, 
                                  cloneCall="strict", 
                                  scale = TRUE, 
                                  exportTable = TRUE)
clonalQuant_output
```

###  Citing scRepertoire

When using `scRepertoire` in your research, please cite the appropriate version of the package.

Citation Details

* **Version 2**: Yang, Q, & Safina, K., Nguyen, K., Tuong, Z.K., & Borcherding, N.  (2025). "scRepertoire 2: Enhanced and efficient toolkit for single-cell immune profiling." *PLoS Computational Biology* https://doi.org/10.1371/journal.pcbi.1012760
* **Version 1**: Borcherding, Nicholas, Nicholas L. Bormann, and Gloria Kraus. "scRepertoire: An R-based toolkit for single-cell immune receptor analysis." *F1000Research* https://doi.org/10.12688/f1000research.22139.2

### Bug Reports/New Features

Your feedback is valuable for improving scRepertoire! If you encounter a bug or have a suggestion for a new feature, please report it.

Submit a [GitHub issue](https://github.com/BorchLab/scRepertoire/issues) - if possible please include a [reproducible example](https://reprex.tidyverse.org/). Alternatively, an example with the internal **scRep_example** and **contig_list** would be extremely helpful.
