---
title: "Frequently Asked Questions"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{Frequently Asked Questions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
p.caption {
  font-size: 0.9em;
}
</style>
 
```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)

suppressMessages(library(scRepertoire))
data("contig_list") 
combined.TCR <- combineTCR(contig_list, 
                           samples = c("P17B", "P17L", "P18B", "P18L", 
                                            "P19B","P19L", "P20B", "P20L"))
```

### Barcode Mismatch

Depending on the pipeline used to generate the single-cell object, there may be inherent mismatches in the barcodes in the single-cell object and the output of ```combineBCR()``` or ```combineTCR()```. In particular, by default, Seurat will amend the suffix of the barcodes with _X, so the barcodes change like:

```
original: ACGTACGTACGTACGT-1
seurat-modified: ACGTACGTACGTACGT-1_1
```

scRepertoire uses the samples in ```combineTCR()``` or ```combineBCR()``` to add a prefix to the barcodes (using the **samples** and/or **ID** parameters):

```
original: ACGTACGTACGTACGT-1
scRepertoire-modified: Sample1_ACGTACGTACGTACGT-1
```

The easiest way to make these compatible is to rename the cell barcodes in the Seurat object by using the `RenameCells()` from the SeuratObject package.

```{r, eval=FALSE}
cell.barcodes <- rownames(seuratObj[[]])
#removing the _1 at the end of the barcodes)
cell.barcodes <- stringr::str_split(cell.barcodes, "_", simplify = TRUE)[,1] 
#adding the prefix of the orig.ident to the barcodes, assuming that is the sample ids
cell.barcodes <- paste0(seuratObj$orig.ident, "_", cell.barcodes) 
seuratObj <- RenameCells(seuratObj, new.names = cell.barcodes)
```

### Adjusting Color Palettes

For all visualizations in scRepertoire, there are 2 ways to adjust the color scheme:  

* Change the **palette** parameter to the desired color scheme. This approach uses the built-in palettes of grDevices and we can access the list of color schemes using ``hcl.pals()``.  
* Add a ggplot layer with a new color scheme.  

```{r tidy = FALSE}
clonalQuant(combined.TCR, 
            cloneCall="strict", 
            chain = "both", 
            scale = TRUE, 
            palette = "Zissou 1")

clonalQuant(combined.TCR, 
            cloneCall="strict", 
            chain = "both", 
            scale = TRUE) + 
  scale_fill_manual(values = hcl.colors(8,"geyser"))
```

### Adjusting Order of Plotting

The order of the group plotting (whether along an axis or in color) can be directly ordered using the **order.by** parameter in functions. Here we can place a vector of strings that will set the exact plotting order. If using **order.by** it is important to make sure your vectors match the **group.by** strings exactly. 

Alternatively, we can set **order.by** = "alphanumeric", if we would like the plots sorted in order of the **group.by** variable.

```{r}
clonalQuant(combined.TCR, 
            cloneCall="strict", 
            chain = "both", 
            scale = TRUE, 
            order.by = c("P17B","P18B","P19B","P20B","P17L","P18L","P19L","P20L"))
```


### Getting Data Used in Plots

Within each of the general analysis functions, there is the ability to export the data frame used to create the visualization. To get the exported values, use **exportTable** = TRUE. It will return the data frame used to make the graph instead of the visual output.  

```{r tidy = FALSE}
clonalQuant_output <- clonalQuant(combined.TCR, 
                                  cloneCall="strict", 
                                  scale = TRUE, 
                                  exportTable = TRUE)
clonalQuant_output
```

###  Citing scRepertoire

We are working on submitting the scRepertoire as a peer review article,

### Bug Reports/New Features

Submit a [GitHub issue](https://github.com/BorchLab/scRepertoire/issues) - if possible please include a [reproducible example](https://reprex.tidyverse.org/). Alternatively, an example with the internal **scRep_example** and **contig_list** would be extremely helpful.
