---
title: "Combining Immcantation and scRepertoire"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{Combining Immcantation and scRepertoire}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include = FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))

knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)
```

# Overview

The Immcantation framework is a suite of R packages designed for the analysis of B cell receptor (BCR) sequencing data. This guide will walk you through the installation and loading of the core packages required for analyzing single-cell BCR data with accompanying interaction with  `scRepertoire`. 

## Installation 

Here is a brief overview of the role each package plays in the Immcantation workflow:

* `alakazam`: This package provides general tools for working with AIRR (Adaptive Immune Receptor Repertoire) data. It includes functions for data import/export, quality control, and basic frequency analysis of V(D)J genes and clonotypes.
* `dowser`: After identifying clones with scoper, dowser is used to perform phylogenetic analysis on the sequences within each clone. This helps in reconstructing the lineage trees that trace the affinity maturation process from the unmutated common ancestor.
* `scoper`: The `scoper` package is used for inferring B cell clones from BCR sequencing data. It implements several methods for partitioning sequences into clonal groups based on shared VJ genes and junction region similarity.
* `shazam`: This package focuses on the analysis of somatic hypermutation (SHM) in immunoglobulin (Ig) sequences. Its functions allow you to quantify mutation frequencies, analyze mutational spectra, and measure the strength and nature of selection pressures.

```{r eval = FALSE}
packages <- c("alakazam", "shazam", "scoper", "dowser")

install.packages(packages)
```

## Citation 


## Loading Packages for the Tutorial

```{r}
library(alakazam)
library(dowser)
library(ggplot2)
library(ggtree)
library(shazam)
library(scoper)
library(scRepertoire)
library(stringr)
library(viridis)
```

# Loading BCR Data

After processing the raw FASTA files with IgBLAST using the provided bash script (Immcantation_Bash.sh), you will have a set of tab-separated value (`.tsv`) files. The bash script automates the use of Change-O's `AssignGenes.py` and `MakeDb.py` tools, which perform V(D)J gene assignment and construct a database file for each sample. The output file we are interested in is `all_contig_igblast_db-pass.tsv`.

Our first task in R is to locate all of these database files, load them, and combine them into a single, clean data frame. The following code block handles this process, performing several essential quality control steps along the way to ensure the data is suitable for downstream analysis.

```{r eval = FALSE}
# -------- Building BCR Data Base -------------------

# Recursively find  files that match the igBlast output
file.list <- list.files(
  path = "./Alignments",
  pattern = "all_contig_igblast_db-pass.tsv",
  recursive = TRUE
)

# Generating Categorical Variables to use Downstream
sample_id <- stringr::str_split(file.list, "/", simplify = TRUE)[, 1]
subject_id <- stringr::str_remove(sample_id, "PBMC_")
subject_id <- stringr::str_remove(subject_id, "_1")
subject_id <- stringr::str_remove(subject_id, "_2")

# Iterative Processing and Filtering of Each Sample ---
lib_data <- lapply(seq_along(file.list), function(x) {
  
  tmp <- read.delim(paste0("./Alignments/", file.list[x]))
  tmp$sample_id <- sample_id[x]
  tmp$subject_id <- subject_id[x]
  tmp$cell_id <- stringr::str_split(tmp$sequence_id, "_", simplify = TRUE)[, 1]
  tmp$cell_id_unique <- paste0(sample_id[x], "_", tmp$cell_id)
  
  
  # Data Cleaning and Filtering Steps 
  if (any(tmp$productive == "TRUE")) {
    tmp <- tmp %>% filter(productive == TRUE)
  } else if (any(tmp$productive == "T")) {
    tmp <- tmp %>% filter(productive == "T")
  }
  
  # Remove cells that have more than one heavy chain contig assigned.
  multi_heavy <- tmp %>%
    filter(locus == "IGH") %>% # Isolate heavy chain records
    count(cell_id) # Count heavy chains per cell
  
  multi_heavy_cells <- multi_heavy %>%
    filter(n > 1) %>%
    pull(cell_id) # Extract the cell IDs as a vector
  
  # Filter out all records (both heavy and light) from these problematic cells.
  tmp <- tmp %>% filter(!cell_id %in% multi_heavy_cells)
  
  # Standardize isotype calls ('c_call').
  tmp$c_call <- stringr::str_split(tmp$c_call, ",", simplify = TRUE)[, 1]
  
  # Remove any records where an isotype could not be determined.
  tmp <- tmp %>% filter(c_call != "")
  
  # Ensure all remaining cells have a paired heavy chain.
  heavy_cells <- tmp %>% filter(locus == "IGH") %>% pull(cell_id)
  light_cells <- tmp %>% filter(locus == "IGK" | locus == "IGL") %>% pull(cell_id)
  no_heavy_cells <- setdiff(light_cells, heavy_cells)
  
  # Remove the orphan light chain records.
  tmp <- tmp %>% filter(!cell_id %in% no_heavy_cells)
  
  return(tmp)
})

# Combining all the data together
bcr_db <- do.call(rbind, lib_data)
```

# Determining a Clonal Assignment Threshold

The core task in B cell repertoire analysis is to group sequences into "clones"—sets of cells that originated from the same initial B cell and underwent affinity maturation. The Immcantation workflow defines clones as sequences that use the same V and J genes and have a high degree of similarity in their junction regions.

But how do we define "high similarity"? We need to determine a specific, objective distance threshold. Sequences with a junction similarity distance below this threshold will be considered part of the same clone. This step uses an automated, data-driven approach to find that threshold.

### Calculate Nearest Neighbor Distances (distToNearest)

The first step is to calculate the Hamming distance (the number of positions at which the characters are different) between the junction sequence of each B cell and its "nearest neighbor." The nearest neighbor is the sequence from a different cell that has the most similar junction.

```{r}
# split by subject id
dist_heavy <- distToNearest(bcr_db, 
                            cellIdColumn = "cell_id_unique",
                            first = FALSE,
                            onlyHeavy = TRUE,
                            fields = "subject_id",
                            nproc = 2, 
                            progress = TRUE)
```

##  Find the Threshold Automatically (`findThreshold`)

The distribution of nearest-neighbor distances is typically bimodal. There is a sharp peak of very small distances (sequences within the same clone) and a broader peak of larger distances (sequences from different clones). The `findThreshold` function from the `scoper` package automatically finds the valley between these two peaks.

```{r}
# find threshold for cloning automatically
threshold_output <- findThreshold(dist_heavy$dist_nearest, 
                                  progress = TRUE,
                                  method = "gmm", 
                                  model = "gamma-norm",
                                  cutoff = "user", 
                                  spc = 0.995)
threshold_auto <- threshold_output@threshold
```

This function fits a Gaussian Mixture Model (`method = "gmm"`) to the distance distribution. It assumes the distribution is a mix of a gamma distribution (for the clonally-related distances) and a normal distribution (for the unrelated distances).

```{r}
ggplot(subset(dist_heavy, !is.na(dist_nearest)),
             aes(x = dist_nearest)) +
  theme_bw() +
  xlab("Hamming distance") + 
  ylab("Count") +
  scale_x_continuous(breaks = seq(0, 1, 0.1)) +
  geom_histogram(color = "white", binwidth = 0.02) +
  geom_vline(xintercept = threshold_auto, lty=2, color='red') +
  facet_grid(subject_id ~ ., scales="free_y") +
  theme(strip.text.y.right = element_text(angle = 0))
```

# Defining and Quantifying Clonal Groups

With the distance threshold established, we can now formally partition our sequences into clonal lineages. This step uses the `scoper` package to apply the threshold and assign a unique clone ID to each sequence. We will then visualize the resulting clone size distributions, which is a fundamental way to characterize an immune repertoire.

## Assigning Sequences to Clones (`hierarchicalClones`)

```{r}
# call clones based on heavy chain
# keep light chain information for subclone calling later
results <- hierarchicalClones(dist_heavy, 
                              cell_id = 'cell_id_unique',
                              threshold = threshold_auto,
                              only_heavy = FALSE,
                              split_light = FALSE, 
                              summarize_clones = TRUE,
                              nproc = 2, 
                              verbose = TRUE)

results_db <- as.data.frame(results)
results_db$sequence_id <- paste0(results_db$sample_id, "_", results_db$sequence_id)
```

## Counting Clone Sizes (`countClones`)

Now that clones are defined, a common first analysis is to quantify their sizes. How many cells belong to each clone?

```{r}
# get clone sizes per subject
clone_sizes <- countClones(results_db,
                           groups=c("locus", "subject_id"))

ggplot(clone_sizes %>% filter(locus=="IGH"), 
       aes(x=seq_count)) +
  geom_bar(fill = "black", color = "black") + 
  geom_text(aes(label=seq_count, y=1), vjust=1.5, size=1) +
  facet_wrap(~subject_id) +
  scale_y_log10() +
  scale_x_sqrt() +
  theme_bw() +
  labs(x="Cells per clone", 
       y="# Clonal lineages")
```

# Germline Reconstruction and Subclone Definition

Now that we have defined our clonal groups based on heavy chain similarity, the next steps are to prepare for mutation analysis and to refine our clone definitions using the paired light chain information available in single-cell data.

This involves two main goals:

* Reconstruct the unmutated germline sequence for each observed BCR. This is the baseline against which we will measure somatic hypermutation.
* Define subclones by identifying groups of cells within a heavy-chain clone that share the same light chain.

## Reconstructing Germline Sequences (`createGermlines`)

```{r}
# Read in IMGT-gapped sequences
references = readIMGT(dir = "~/share/germlines/imgt/human/vdj")
results_db <- createGermlines(results_db, 
                              reference = references)

# Data cleaning: Some downstream functions do not tolerate NA values in columns
results_db$c_call[is.na(results_db$c_call)] <- "NA"
```

## Defining Subclones by Light Chain (`resolveLightChains`)

A key advantage of single-cell data is having paired heavy and light chain information. While we defined our clones using only the heavy chain, cells within that clone may use different light chains. These represent distinct "subclones." The `resolveLightChains` function from `scoper` identifies these groupings.

```{r}
# split clones into subclones by light chain
subclones_db <- resolveLightChains(data=results_db,
                                   cell = "cell_id_unique", 
                                   id = "sequence_id", 
                                   nproc = 2)


# General Clean Up
subclones_db$v_gene <- str_split(subclones_db$v_call, "[*]", simplify = TRUE)[,1]
subclones_db$j_gene <- str_split(subclones_db$j_call, "[*]", simplify = TRUE)[,1]
subclones_db$c_gene <- str_split(subclones_db$c_call, "[*]", simplify = TRUE)[,1]
subclones_db <- subclones_db %>%
  mutate(chain = ifelse(locus == "IGH", "heavy", "light"))

subclones_db %>% 
  group_by(subject_id, chain, c_gene) %>% 
  tally() %>%
  ggplot(aes(x=subject_id, y=n)) +
  geom_bar(aes(fill=factor(c_gene)), 
           stat="identity", position="stack", color = "black", lwd = 0.25) +
  facet_wrap(~chain) +
  scale_fill_viridis(option = "B", discrete = TRUE, name="Isotype") + 
  labs(x="Subject", y="# BCRs") +
  theme_bw()
```

# Preparing Data for Lineage Tree Analysis

The ultimate goal of many BCR repertoire analyses is to reconstruct the evolutionary history of each B cell clone. This is done by building phylogenetic trees that trace the path of somatic hypermutation from the unmutated germline ancestor to the observed, mature B cells.

The `dowser` package is used for this phylogenetic analysis, but it requires the data to be in a specific format: a `ScoperClones` object. This step uses the `formatClones` function to perform some final data cleaning and to create this specialized object.

```{r}
# Remove any rows where junction_length is NA, as this is a critical field.
subclones_db <- filter(subclones_db, !is.na(junction_length))

# Replace NA values in key annotation columns with the string "NA".
subclones_db$c_gene[is.na(subclones_db$c_gene)] <- "NA"
subclones_db$v_gene[is.na(subclones_db$v_gene)] <- "NA"
subclones_db$j_gene[is.na(subclones_db$j_gene)] <- "NA"
subclones_db$vj_gene[is.na(subclones_db$vj_gene)] <- "NA"
subclones_db$tissue <- str_split(subclones_db$cell_id_unique, "_", simplify = TRUE)[,2]

clones_hl <- formatClones(subclones_db,
                          cell = 'cell_id_unique',
                          chain = "HL", 
                          split_light = FALSE,
                          traits = c("subject_id", "c_gene"),
                          columns = c("c_gene", "v_gene", "j_gene", "vj_gene", "subject_id", "tissue"),
                          minseq = 2, 
                          verbose = FALSE)
```

# Building and Visualizing Clonal Trees

The culmination of the `Immcantation` workflow is the reconstruction and visualization of clonal lineage trees. These trees map the evolutionary relationships between sequences in a clone, starting from the unmutated germline ancestor and tracing the branches of somatic hypermutation and class-switching that occur during affinity maturation. 

First, we use `getTrees()` function to build a maximum parsimony tree for each clone, and then we scale the branches to represent the number of mutations. We will then use `collapseNodes()` to cleanup the trees and `scaleBranches()` to enhanced the visual comparison of the tree nodes. 

```{r}
clones_hl$num_cells <- sapply(clones_hl$data, function(x) sum(x@data$collapse_count))

# --- Build trees for each clone ---
trees = getTrees(clones_hl)
trees <- collapseNodes(trees)
trees_m = scaleBranches(trees, edge_type="mutations")
```
## Visualizing Clonal Trees

```{r}
# --- Generate and Customize Tree Plots ---

# Define a color palette for isotypes for consistent plotting
isotypes <- c("IGHD", "IGHM", "IGHG1", "IGHG2", "IGHG3", "IGHG4", "IGHA1", "IGHA2", "Germline")
col_pal <- c(viridis_pal(option = "B")(8), "grey")
names(col_pal) <- isotypes

# plotTrees generates a list of ggplot objects, one for each tree.
plots_m <- plotTrees(trees_m, scale=1) 
names(plots_m) <- trees_m$clone_id

# --- Example: Plotting a Single Interesting Tree ---
idx <- trees_m %>% 
  slice_max(order_by = seqs, n = 3) %>% 
  pull(clone_id)
example_tree_plot <- plots_m[[idx]]

plot_data <- example_tree_plot$data
plot_data$tissue <- str_split(plot_data$cell_id_unique, "_", simplify = TRUE)[,2]
plot_data$tissue <- ifelse(is.na(plot_data$tissue), "Germline", plot_data$tissue) # Handle germline node
example_tree_plot$data <- plot_data
max_x <- max(example_tree_plot$data$x) # Find x-axis limit

example_tree_plot + 
    labs(
        title = paste("Lineage Tree for Clone:", example_tree_plot$data$clone_id[1]),
        subtitle = "Branches scaled by number of mutations"
    ) +
    geom_tippoint(aes(color = c_gene, 
                      shape = tissue, 
                      size = collapse_count), 
                  alpha = 0.8) + 
    
    scale_color_manual(values = col_pal, name = "Isotype") +
    scale_shape_manual(values = c("Germline"=1, "PT"=16, "LN"=17), name = "Tissue") +
    scale_size(name = "# Cells", range = c(2, 6)) +
    xlim(0, max_x + (max_x * 0.5)) +
    theme(
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "italic"),
        legend.position = "right"
    )

```

# Quantifying and Analyzing Somatic Hypermutation

A hallmark of the adaptive immune response is somatic hypermutation (SHM), the process by which B cells introduce mutations into their antibody genes to improve antigen binding. Quantifying the frequency, location, and type of these mutations is a primary goal of BCR repertoire analysis. This step uses the shazam package to calculate SHM and explore its patterns.

## Calculate Somatic Hypermutation (`observedMutations()`)

The `observedMutations()` function is the workhorse for all SHM analysis. It compares each observed sequence to its reconstructed germline sequence (which we created in Step 4) and tallies the mutations. The function is called multiple times to calculate different metrics. It compares the `sequence_alignment` (observed sequence) to the `germline_alignment_d_mask` (unmutated ancestor).

Key Parameter(s) for `observedMutations()`

* `regionDefinition = IMGT_VDJ`: This tells the function to use the standard IMGT definitions to identify Framework Regions (FWRs) and Complementarity-Determining Regions (CDRs). CDRs are the hypervariable loops that directly contact the antigen, while FWRs provide the structural scaffold.
* `frequency = TRUE`: Calculates the mutation rate (mutations per 100 base pairs). If FALSE, it calculates the raw count of mutations.
* `combine = TRUE`: This calculates a single mutation value across the entire VDJ segment. If FALSE, it calculates mutations separately for each sub-region (e.g., mu_freq_cdr_s for silent mutations in the CDR, mu_freq_fwr_r for replacement mutations in the FWR, etc.).

After running these commands, your shm_db data frame will be populated with many new columns (e.g., mu_freq, mu_count, mu_freq_cdr_s, mu_freq_cdr_r) that quantify SHM in great detail.

```{r}
shm_db <- observedMutations(subclones_db,
                            sequenceColumn = "sequence_alignment",
                            germlineColumn = "germline_alignment_d_mask",
                            regionDefinition = IMGT_VDJ,
                            frequency = FALSE,
                            combine = FALSE,
                            nproc = 2)

test <- observedMutations(subclones_db, 
                          sequenceColumn="sequence_alignment",
                          germlineColumn="germline_alignment_d_mask",
                          regionDefinition=NULL,
                          frequency=FALSE, 
                          nproc=2)

shm_db <- observedMutations(shm_db,
                            sequenceColumn = "sequence_alignment",
                            germlineColumn = "germline_alignment_d_mask",
                            regionDefinition = IMGT_VDJ,
                            frequency = FALSE,
                            combine = TRUE,
                            nproc = 2)

# calculate SHM frequency in the VDJ regions
shm_db <- observedMutations(shm_db,
                            sequenceColumn = "sequence_alignment",
                            germlineColumn = "germline_alignment_d_mask",
                            regionDefinition = IMGT_VDJ,
                            frequency = TRUE,
                            combine = FALSE,
                            nproc = 2)

shm_db <- observedMutations(shm_db,
                            sequenceColumn = "sequence_alignment",
                            germlineColumn = "germline_alignment_d_mask",
                            regionDefinition = IMGT_VDJ,
                            frequency = TRUE,
                            combine = TRUE,
                            nproc = 2)
```
### Visualizing Replacement and Silent Mutations in CDRs

A key way to infer positive or negative selection is to compare the rate of replacement (**R**) mutations (which change the amino acid sequence) to silent (**S**) mutations (which do not). High levels of R mutations in the antigen-binding CDRs are a classic sign of positive selection for improved binding.

```{r}
# Clean up isotype calls for plotting
shm_db$c_call <- str_split(shm_db$c_call, "[*]", simplify = TRUE)[,1]

# Plot silent mutation frequency in the CDRs
shm_db %>%
  filter(locus == "IGH" & c_call != "NA") %>%
  ggplot(shm_db, aes(x = subject_id, y=mu_freq_cdr_s, fill=c_call)) +
          geom_boxplot(outlier.alpha = 0) + 
          labs(title = "CDR silent mutations", 
               x = "Isotype", y = "Mutation frequency") +
          scale_fill_viridis(option = "H", discrete = TRUE) +
          facet_grid(.~c_call, scales="free_y") + 
     theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

# Plot replacement mutation frequency in the CDRs
shm_db %>%
  filter(locus == "IGH" & c_call != "NA") %>%
  ggplot(aes(x = subject_id, y=mu_freq_cdr_r, fill=c_call)) +
          geom_boxplot(outlier.alpha = 0) + 
          labs(title = "CDR replacement mutations", 
               x = "Isotype", y = "Mutation frequency") +
          scale_fill_viridis(option = "H", discrete = TRUE) +
          facet_grid(.~c_call, scales="free_y") + 
     theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

This boxplot displays the frequency of replacement mutations specifically within the CDRs of heavy chains. It is faceted by isotype (`c_call`), allowing you to compare SHM levels in naive (IgM/IgD) versus mature (IgG/IgA) B cells. You would typically expect to see higher R mutation frequencies in class-switched isotypes, indicating that these clones have undergone selection. 

### Visualizing Overall SHM Distribution

Finally, we can look at the overall distribution of SHM across all clones to get a global picture of the repertoire's maturity.

```{r}
# First, calculate the mean mutation frequency for each clone
mut_freq_clone <- shm_db %>%
  filter(locus == "IGH") %>%
  group_by(clone_id, subject_id, c_call) %>%
  summarise(mean_mut_freq = mean(mu_freq))

# Order isotypes for consistent plotting
isotypes = c("IGHD", "IGHM", "IGHA1", "IGHA2", "IGHE", "IGHG1", "IGHG2", "IGHG3", "IGHG4")
mut_freq_clone$c_call <- factor(mut_freq_clone$c_call, levels=isotypes)

# Create a density plot of mutation frequencies
ggplot(mut_freq_clone %>% filter(c_call != "NA"), 
       aes(mean_mut_freq, fill = c_call)) +
  geom_density(alpha=0.5) +
  facet_grid(c_call~subject_id, scales="free") +
  labs(x="Mean mutation frequency per clone", y="Density",
       title="Somatic Hypermutation Rate by Isotype and Subject") +
  theme_bw()
```

This series of density plots shows the distribution of SHM frequencies for each isotype within each subject. The x-axis represents the mutation frequency. A peak near zero indicates a large population of naive or unmutated cells (such as expected for IGHM and IGHD). In a mature immune response, you would expect to see the peaks for IgG and IgA shifted to the right, indicating higher overall mutation levels. 