---
title: "Downstream Analysis with immunarch"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{Downstream Analysis with immunarch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include = FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))

knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)

suppressMessages(library(scRepertoire))
suppressMessages(library(tcrpheno))
suppressMessages(library(Seurat))

data("contig_list") 
combined.TCR <- combineTCR(contig_list, 
                           samples = c("P17B", "P17L", "P18B", "P18L", 
                                            "P19B","P19L", "P20B", "P20L"))
```

# Overview

The `scRepertoire` package provides robust tools for the initial processing, filtering, and combining of single-cell immune receptor sequencing data. In addition to `scRepertoire`, users have used the `immunarch` package, which offers a powerful and comprehensive suite of immune profiling functions.

This vignette demonstrates the seamless integration between `scRepertoire` and `immunarch`. We will use the `exportClones()` function with the `format = "immunarch"` option to generate a compatible data object, which can then be directly used for downstream analysis and visualization with `immunarch`.

## Citation 

If using *immunarch*, please cite the [package](https://immunarch.com/authors.html#citation)

```
@Manual{,
  title = {immunarch: Bioinformatics Analysis of T-Cell and B-Cell Immune Repertoires},
  author = {Vadim I. Nazarov and Vasily O. Tsvetkov and Siarhei Fiadziushchanka and Eugene Rumynskiy and Aleksandr A. Popov and Ivan Balashov and Maria Samokhina},
  year = {2023},
  note = {https://immunarch.com/, https://github.com/immunomind/immunarch},
}
```

# Setup

First, we need to load the necessary libraries and prepare the initial data using `scRepertoire`. We will use the built-in `contig_list` example data.

```{r LoadLibraries}
suppressMessages(library(scRepertoire))
suppressMessages(library(immunarch))
suppressMessages(library(ggplot2))

# Load example data from scRepertoire
data("contig_list") 

# Combine contigs into a single list
combined <- combineTCR(contig_list, 
                       samples = c("P17B", "P17L", "P18B", "P18L", 
                                   "P19B", "P19L", "P20B", "P20L"))
```

# Exporting for `immunarch`

The `exportClones()` function can now format the `scRepertoire` object into a list that `immunarch` can immediately use. This list contains two key components:
* **`$data`**: A list of data frames, where each data frame is a repertoire from a single sample.
* **`$meta`**: A metadata data frame describing the samples.

```{r ExportClones}
# Export the data with write.file = FALSE to get the R object
immunarch_data <- exportClones(combined, 
                               format = "immunarch", 
                               write.file = FALSE)

# We can inspect the structure of the output
str(immunarch_data, max.level = 2)
```

As you can see, the output is already in the format required by `immunarch`. Now we can proceed with downstream analysis.

# Basic Analysis with `immunarch`

Let's perform a few common analyses to demonstrate the workflow. For visualization purposes, we will add a "SampleType" column to the metadata to group the samples by their origin (B for blood, L for lymph node).

```{r AddMetadata}
# Add a grouping variable to the metadata
immunarch_data$meta$SampleType <- substr(immunarch_data$meta$Sample, 4, 4)
head(immunarch_data$meta)
```

## Repertoire Overlap

We can measure the similarity between repertoires by calculating the number of shared clonotypes (public clonotypes) using `repOverlap()`. The `vis()` function can then generate a heatmap to visualize the pairwise overlap.

```{r OverlapAnalysis}
# Calculate overlap using the number of public clonotypes
imm_ov <- repOverlap(immunarch_data$data, .method = "public", .verbose = FALSE)

# Visualize the overlap as a heatmap
vis(imm_ov)
```

## Clonal Homeostasis

Next, let's assess the clonal space homeostasis, which is the proportion of the repertoire occupied by clones of different sizes (e.g., Small, Medium, Large). The `repClonality()` function calculates this, and `vis()` creates a bar plot.

```{r ClonalityAnalysis}
# Calculate clonal space homeostasis
imm_hom <- repClonality(immunarch_data$data,
                        .method = "homeo",
                        .clone.types = c(Small = 0.0001, Medium = 0.001, 
                                         Large = 0.01, Hyperexpanded = 1))

# Visualize homeostasis, grouped by sample type
vis(imm_hom, .by = "SampleType", .meta = immunarch_data$meta)
```

## Gene Usage

`immunarch` provides powerful tools to analyze and visualize the usage of V, D, and J genes. Here, we'll compute the usage of TRBV genes and visualize their distribution across samples.

```{r GeneUsage}
# Compute TRBV gene usage
imm_gu <- geneUsage(immunarch_data$data[1], "hs.trbv", .norm = TRUE)

# Visualize gene usage as a heatmap
vis(imm_gu, .grid = TRUE, .title = "TRBV Gene Usage")
```

We can also visualize the gene usage grouped by our metadata variable.

```{r GeneUsageGrouped}
imm_gu <- geneUsage(immunarch_data$data, "hs.trbv", .norm = T)
imm_gu_js <- geneUsageAnalysis(imm_gu, .method = "js", .verbose = F)
imm_gu_cor <- geneUsageAnalysis(imm_gu, .method = "cor", .verbose = F)

p1 <- vis(imm_gu_js, .title = "Gene usage JS-divergence", .leg.title = "JS", .text.size = 1.5)
p2 <- vis(imm_gu_cor, .title = "Gene usage correlation", .leg.title = "Cor", .text.size = 1.5)

p1 + p2
```

This concludes the vignette on integrating `scRepertoire` with `immunarch`. By using the `exportClones()` function, you can easily leverage the extensive analytical and visualization capabilities of `immunarch` for your single-cell TCR sequencing data.

