---
title: "Additional Processing Steps"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{Additional Processing Steps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
p.caption {
  font-size: 0.9em;
}
</style>
 
```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)

suppressMessages(library(scRepertoire))
data("contig_list") 
combined.TCR <- combineTCR(contig_list, 
                           samples = c("P17B", "P17L", "P18B", "P18L", 
                                       "P19B","P19L", "P20B", "P20L"))
```

## addVariable

What if there are more variables to add than just sample and ID? We can add them by using the `addVariable()` function. All we need is the **variable.name** of the variable you'd like to add and the specific character or numeric values (**variables**). As an example, here we add the **Type** in which the samples were processed and sequenced. 

```{r tidy = FALSE}
combined.TCR <- addVariable(combined.TCR, 
                            variable.name = "Type", 
                            variables = rep(c("B", "L"), 4))

head(combined.TCR[[1]])
```

## subsetClones

Likewise, we can remove specific list elements after `combineTCR()` using the `subsetClones()` function. In order to subset, we need to identify the vector we would like to use for subsetting (**name**) and the variable values to subset (**variables**). Below, we isolate just the 2 sequencing results from P18L and P18B. 

```{r, tidy = FALSE}
subset1 <- subsetClones(combined.TCR, 
                        name = "sample", 
                        variables = c("P18L", "P18B"))

head(subset1[[1]])
```

Alternatively, we can also just select the list elements after `combineTCR()` or `combineBCR()`.

```{r}
subset2 <- combined.TCR[c(3,4)]
head(subset2[[1]])
```

## exportClones  

After assigning the clone by barcode, we can export the paired clonotypes using `exportClones()` to save for later use or to use in other pipelines. 

**format**  

* "paired" - Export the paired sequences **(default)**.  
* "airr" - Export data in an AIRR-compliant format.  
* "TCRMatch" - Export TCRB chain information. 
    
**write.file**    

* TRUE, save the file.  
* FALSE, return a data.frame.  

**dir**  
directory location to save the csv  

**file.name**  
the csv file name  

```{r, eval = FALSE, tidy = FALSE}
exportClones(combined, 
             write.file = TRUE,
             dir = "~/Documents/MyExperiment/Sample1/"
             file.name = "clones.csv")
```

## annotateInvariant

The ```annotateInvariant()``` function enables the identification of mucosal-associated invariant T (**MAIT**) cells and invariant natural killer T (**iNKT**) cells in single-cell sequencing datasets. These specialized T-cell subsets are defined by their characteristic TCR usage, making them distinguishable within single-cell immune profiling data. The function extracts TCR chain information from the provided single-cell dataset and evaluates it against known invariant TCR criteria for either MAIT or iNKT cells. Each cell is assigned a score indicating the presence (1) or absence (0) of the specified invariant T-cell population.

* input.data – A data object from ```combineTCR()``` or ```combineExpression()```.
* type – Character string specifying the type of invariant T cell to annotate ("MAIT" or "iNKT").
* species – Character string specifying the species ("mouse" or "human").

```{r, eval = FALSE, tidy = FALSE}
combined <- annotateInvariant(combined, 
                              type = "MAIT", 
                              species = "human")

combined <- annotateInvariant(combined, 
                              type = "iNKT", 
                              species = "human")
```