---
title: "Summarizing Repertoires"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{Summarizing Repertoires}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
p.caption {
  font-size: 0.9em;
}
</style>
 
```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)

suppressMessages(library(scRepertoire))
data("contig_list") 
combined.TCR <- combineTCR(contig_list, 
                           samples = c("P17B", "P17L", "P18B", "P18L", 
                                            "P19B","P19L", "P20B", "P20L"))
```

## percentAA

Quantify the proportion of amino acids along the cdr3 sequence with `percentAA()`. By default, the function will pad the sequences with NAs up to the maximum of **aa.length**. Sequences longer than **aa.length** will be removed before visualization **(default aa.length = 20)**. 

```{r, message = FALSE, tidy = FALSE}
percentAA(combined.TCR, 
          chain = "TRB", 
          aa.length = 20)
```

## positionalEntropy

We can also quantify the level of entropy/diversity across amino acid residues along the cdr3 sequence. `positionalEntropy()` combines the quantification by residue of ``percentAA()`` with the diversity calls in ```clonalDiversity()```. 

**method**  

* "shannon" - Shannon Diversity  
* "inv.simpson" - Inverse Simpson Diversity 
* "norm.entropy" - Normalized Entropy  

```{r tidy = FALSE}
positionalEntropy(combined.TCR, 
                  chain = "TRB", 
                  aa.length = 20)
```
## positionalProperty

Like ```positionalEntropy()```, we can also examine a series of amino acid properties along the cdr3 sequences using ```positionalProperty()```. Important differences from the above function for ```positionalProperty()``` is dropping NA values as they would void the mean calculation. ```positionalProperty()``` also display a ribbon with the 95% confidence interval surrounding the mean value for the selected properties.

**method**  

* "Atchley" - Atchley Factors 
* "Kidera" - Kidera Factors 
* "stScales" - stScales Vectors
* "tScales" - tScales Vectors
* "VHSE" - Vectors of Hydrophobic, Steric, and Electronic properties

```{r tidy = FALSE}
positionalProperty(combined.TCR[c(1,2)], 
                  chain = "TRB", 
                  aa.length = 20, 
                  method = "Atchley") + 
  scale_color_manual(values = hcl.colors(5, "inferno")[c(2,4)])
```
## vizGenes

A visualization of the relative usage of genes of the TCR or BCR, using `vizGenes()`. There is some functional crossover between `vizGenes()` and two functions below called  `percentGenes()` and `percentVJ()`. But `vizGenes()` is more adaptable to allow for comparisons across chains, scaling, etc. 

**x.axis**  

* Gene segment to separate the x-axis, such as "TRAV", "TRBD", "IGKJ"

**y.axis** 

* Gene segment or categorical variable to visualize along the y-axis

**plot**  

* "barplot" for a bar chart  
* "heatmap" for a heatmap  

**scale**  

* TRUE to scale the graph by number of genes per sample  
* FALSE to report raw numbers  

**order**  

* "gene" to order by gene name  
* "variance" to order by variance between the separate variable categories  

```{r tidy = FALSE}
vizGenes(combined.TCR, 
         x.axis = "TRBV",
         y.axis = NULL,
         plot = "barplot",  
         scale = TRUE)
```

`vizGenes()` can also be used to look at the usage of genes in a single chain. For example, say we are interested in the difference in TRB V and J usage between lung and peripheral blood samples - we can easily take a look at this using the following code:

```{r tidy = FALSE}
#Peripheral Blood
vizGenes(combined.TCR[c(1,3,5,7)], 
         x.axis = "TRBV",
         y.axis = "TRBJ",
         plot = "heatmap",  
         scale = TRUE)

#Lung
vizGenes(combined.TCR[c(2,4,6,8)], 
         x.axis = "TRBV",
         y.axis = "TRBJ",
         plot = "heatmap",  
         scale = TRUE)
```

For the P17 patient samples, what if we are interested in chain pairings, we can look at TRBV and TRAV at the same time using them as inputs to **x.axis** and **y.axis**. 

```{r tidy = FALSE}
vizGenes(combined.TCR[c(1,2)], 
         x.axis = "TRBV",
         y.axis = "TRAV",
         plot = "heatmap", 
         scale = FALSE)
```


## percentGenes

Quantify the proportion of V or J gene usage with `percentGenes()`. Like `percentAA()`, we select the chain of interest and then indicate the gene of interest with the **gene** parameter. Two major limitations of `percentGenes()` are, 1) the function quantifies only V or J genes, and 2) the quantification of the genes are limited to all the V or J genes seen across the samples, not all possible V or J genes. 

```{r tidy = FALSE}
percentGenes(combined.TCR, 
             chain = "TRB", 
             gene = "Vgene")
```

We can also use the output `percentGenes()` for dimensional reduction to summarise the gene usage by sample. This can be done with a simple principal component analysis (below) or even more complex reductions. 

```{r tidy = FALSE}
df.genes <- percentGenes(combined.TCR, 
                         chain = "TRB", 
                         gene = "Vgene", 
                         exportTable = TRUE)

#Performing PCA
pc <- prcomp(df.genes)

#Getting data frame to plot from
df <- as.data.frame(cbind(pc$x[,1:2], rownames(df.genes)))
df$PC1 <- as.numeric(df$PC1)
df$PC2 <- as.numeric(df$PC2)

#Plotting
ggplot(df, aes(x = PC1, y = PC2)) + 
  geom_point(aes(fill =df[,3]), shape = 21, size = 5) + 
  guides(fill=guide_legend(title="Samples")) + 
  scale_fill_manual(values = hcl.colors(nrow(df), "inferno")) + 
  theme_classic() 
```

## percentVJ

Quantify the proportion of V and J gene usage with `percentVJ()`. Like `percentGenes()`, this function will quantify the percentage of V and J paired together across individual repertoires. The output can be visualized using a heatmap or as input for further dimensional reduction.

```{r tidy = FALSE}
percentVJ(combined.TCR, 
          chain = "TRB")

df.genes <- percentVJ(combined.TCR, 
                      chain = "TRB", 
                      exportTable = TRUE)

#Performing PCA
pc <- prcomp(df.genes)

#Getting data frame to plot from
df <- as.data.frame(cbind(pc$x[,1:2], rownames(df.genes))) 
df$PC1 <- as.numeric(df$PC1)
df$PC2 <- as.numeric(df$PC2)

#Plotting
ggplot(df, aes(x = PC1, y = PC2)) + 
  geom_point(aes(fill =df[,3]), shape = 21, size = 5) + 
  guides(fill=guide_legend(title="Samples")) + 
  scale_fill_manual(values = hcl.colors(nrow(df), "inferno")) + 
  theme_classic() 
```

## percentKmer

Another quantification of the composition of the CDR3 sequence is to define motifs by sliding across the amino acid or nucleotide sequences at set intervals resulting in substrings or kmers. 

**motif.length**  

* Numerical value for the length of the kmer.

**top.motifs**  

* Display the most variable genes determined by mean absolute deviation (MAD).

```{r tidy = FALSE}
percentKmer(combined.TCR, 
            cloneCall = "aa",
            chain = "TRB", 
            motif.length = 3, 
            top.motifs = 25)

percentKmer(combined.TCR, 
            cloneCall = "nt",
            chain = "TRB", 
            motif.length = 3, 
            top.motifs = 25)
```