---
title: "Visualizations for Single-Cell Objects"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{Visualizations for Single-Cell Objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
p.caption {
  font-size: 0.9em;
}
</style>
 
```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)

suppressMessages(library(scRepertoire))
suppressMessages(library(Seurat))
data("contig_list") 
combined.TCR <- combineTCR(contig_list, 
                           samples = c("P17B", "P17L", "P18B", "P18L", 
                                            "P19B","P19L", "P20B", "P20L"))
googledrive::drive_download("https://drive.google.com/file/d/1_YuRraDyg8UgF3oasjF0-jgPnwox-B24/view?usp=share_link", overwrite = TRUE)

scRep_example <- readRDS("scRep_example_full.rds")

scRep_example <- combineExpression(combined.TCR, 
                                   scRep_example, 
                                   cloneCall="gene", 
                                   group.by = "sample")

#Adding patient information
scRep_example$Patient <- substr(scRep_example$orig.ident, 1,3)

#Adding type information
scRep_example$Type <- substr(scRep_example$orig.ident, 4,4)

#Defining colors
colorblind_vector <- hcl.colors(n=7, palette = "inferno", fixup = TRUE)
```

## clonalOverlay 

Using the dimensional reduction graphs as a reference, we can also generate an overlay of the position of clonally-expanded cells using `clonalOverlay()`. Select the **reduction** for the visualization, default is "PCA" and the **freq.cutpoint** or lowest clonal frequency or proportion to generate the contour plot. 

We can modify the contours by selecting the number of **bins** or the number of contours drawn. `clonalOverlay()` can be used to look across all cells or faceted by a meta data variable using **facet.by**. The overall dimensional reduction will be maintained as we facet, while the contour plots will adjust based on the **facet.by** variable. The coloring of the dot plot is based on the active identity of the single-cell object. 

This visualization was authored by Dr. Francesco Mazziotta from Johns Hopkins and inspired by Drs. Carmona and Andreatta and their work with [ProjectTIL](https://github.com/carmonalab/ProjecTILs), which is a great pipeline to annotated T cell subtypes.

```{r tidy = FALSE}
clonalOverlay(scRep_example, 
              reduction = "umap", 
              freq.cutpoint = 1, 
              bins = 10, 
              facet.by = "Patient") + 
              guides(color = "none")
```

## clonalNetwork 

Similar to `clonalOverlay()`, we can look at the network interaction of clonotypes shared between clusters along the single-cell dimensional reduction using `clonalNetwork()`. This function shows the relative proportion of clones from the starting node, with the ending node indicated by the arrow. 

Filtering for clones can be accomplished using 3 methods:  

**filter.clones**  
* Select a number to isolate the clones comprising the top X number of cells (filter.clones = 2000)  
* Select "min" to make sure all groups are scaled to the size of the minimum group

**filter.identity**  
* For the identity chosen to visualize, show the to and from network connections for a single group  

**filter.proportion**
* Remove clones that comprise less than a certain proportion of clones in groups.
                          

```{r tidy = FALSE}
#ggraph needs to be loaded due to issues with ggplot
library(ggraph)

#No Identity filter
clonalNetwork(scRep_example, 
              reduction = "umap", 
              group.by = "seurat_clusters",
              filter.clones = NULL,
              filter.identity = NULL,
              cloneCall = "aa")

#Examining Cluster 2 only
clonalNetwork(scRep_example, 
              reduction = "umap", 
              group.by = "seurat_clusters",
              filter.identity = 3,
              cloneCall = "aa")
```

We can also use **exportClones** to quickly get clones that are shared across multiple identity groups, along with the total number of clones in the data set.

```{r tidy = FALSE}
shared.clones <- clonalNetwork(scRep_example, 
                               reduction = "umap", 
                               group.by = "seurat_clusters",
                               cloneCall = "aa", 
                               exportClones = TRUE)
head(shared.clones)
```

## highlightClones

We can also look at the clonotypes by calling specific sequences in the `highlightClonotypes()` below. In order to highlight the clonotypes, we first need to use the **cloneCall**, the type of sequence we will be using, and then the specific sequences themselves using **sequence**. Below, we can see the steps to highlight the most prominent sequences *CAERGSGGSYIPTF_CASSDPSGRQGPRWDTQYF*, and *CARKVRDSSYKLIF_CASSDSGYNEQFF*.

```{r tidy = FALSE}
scRep_example <- highlightClones(scRep_example, 
                    cloneCall= "aa", 
                    sequence = c("CAERGSGGSYIPTF_CASSDPSGRQGPRWDTQYF", 
                                 "CARKVRDSSYKLIF_CASSDSGYNEQFF"))

Seurat::DimPlot(scRep_example, group.by = "highlight") + 
  ggplot2::theme(plot.title = element_blank())
```

## clonalOccupy

We can also look at the count of cells by cluster assigned into specific frequency ranges by using the `clonalOccupy()` function and selecting the **x.axis** to display cluster or other variables in the meta data of the single cell object.

+ **proportion** can be used to look at relative level groupings  
+ **label** will still return the absolute number of clonotypes  

```{r tidy = FALSE}
clonalOccupy(scRep_example, 
                     x.axis = "seurat_clusters")

clonalOccupy(scRep_example, 
                     x.axis = "ident", 
                     proportion = TRUE, 
                     label = FALSE)
```

## alluvialClones

After the metadata has been modified, we can look at clones across multiple categories using the `alluvialClones()` function. We are able to use the plots to examine the interchange of categorical variables. Because this function will produce a graph with each clonotype arranged by called stratification, this will take some time depending on the size of the repertoire.

To understand the basic concepts of this graphing method, I'd highly recommend reading [this post](https://cran.r-project.org/web/packages/ggalluvial/vignettes/ggalluvial.html).  

```{r tidy = FALSE}
alluvialClones(scRep_example, 
               cloneCall = "aa", 
               y.axes = c("Patient", "ident", "Type"), 
               color = c("CVVSDNTGGFKTIF_CASSVRRERANTGELFF", "NA_CASSVRRERANTGELFF")) + 
    scale_fill_manual(values = c("grey", colorblind_vector[3]))


alluvialClones(scRep_example, 
                   cloneCall = "gene", 
                   y.axes = c("Patient", "ident", "Type"), 
                   color = "ident") 
```

## getCirclize

Like alluvial graphs, we can also visualize the interconnection of clusters using the chord diagrams from the [circlize R package](https://jokergoo.github.io/circlize_book/book/). 

The first step is getting the data frame output to feed into the `chordDiagram()` function in circlize, which can be done using `getCirclize()`. 

This will calculate the relative number of clonotypes shared based on the **group.by** variable using the product of `combineExpression()`. It is important to note `getCirclize()` will create a matrix the size of the **group.by** variable and then simplify into instructions to be read by the circlize R package. The output is the total number of unique and shared clones by the **group.by** variable. 

```{r tidy = FALSE}
library(circlize)
library(scales)

circles <- getCirclize(scRep_example, 
                       group.by = "seurat_clusters")

#Just assigning the normal colors to each cluster
grid.cols <- hue_pal()(length(unique(scRep_example$seurat_clusters)))
names(grid.cols) <- unique(scRep_example$seurat_clusters)

#Graphing the chord diagram
chordDiagram(circles, self.link = 1, grid.col = grid.cols)
```