---
title: "Using Trex for TCRs"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{Using Trex for TCRs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
***

<style>
p.caption {
  font-size: 0.9em;
}
</style>
 
```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)

suppressMessages(library(scRepertoire))
suppressMessages(library(Trex))
suppressMessages(library(Seurat))
suppressMessages(library(ggplot2))
suppressMessages(library(viridis))
suppressMessages(library(dplyr))
suppressMessages(library(reticulate))
use_condaenv(condaenv = "r-reticulate", required = TRUE)

data("contig_list") 
combined.TCR <- combineTCR(contig_list, 
                           samples = c("P17B", "P17L", "P18B", "P18L", 
                                            "P19B","P19L", "P20B", "P20L"))

googledrive::drive_download("https://drive.google.com/file/d/1_YuRraDyg8UgF3oasjF0-jgPnwox-B24/view?usp=share_link", overwrite = TRUE)

scRep_example <- readRDS("scRep_example_full.rds")
scRep_example <- scRep_example %>%
                     ScaleData(verbose = FALSE) %>%
                     FindVariableFeatures(verbose = FALSE) %>%
                     RunPCA(verbose = FALSE)


scRep_example <- combineExpression(combined.TCR, 
                                   scRep_example, 
                                   cloneCall="gene", 
                                   group.by = "sample")

#Adding patient information
scRep_example$Patient <- substr(scRep_example$orig.ident, 1,3)

#Adding type information
scRep_example$Type <- substr(scRep_example$orig.ident, 4,4)

#Defining colors
colorblind_vector <- hcl.colors(n=7, palette = "inferno", fixup = TRUE)
```


The idea behind Trex is to combine TCR CDR3 amino acid information with phenotypic RNA/protein data to direct the use of single-cell sequencing towards antigen-specific discoveries. This is a growing field - specifically [TESSA](https://github.com/jcao89757/TESSA) uses amino acid characteristics and autoencoder as a means to get a dimensional reduction. Another option is [CoNGA](https://github.com/phbradley/conga), which produces an embedding using TCR and RNA with python. Trex was designed to make a customizable approach to this combined approach using R.

More infromation is available at the [Trex GitHub Repo](https://github.com/ncborcherding/Trex).

## Installation

```{r, eval = FALSE}
devtools::install_github("ncborcherding/Trex")
```


## maTrex Function

Trex has 2 major functions - the first being ```maTrex()```, which is the backbone of the algorithm and returns the encoded values based on the selection of variables. Unlike ```runTrex()``` below, ```maTrex()``` does not filter the input for only T cells with attached TCR data. In addition, ```maTrex()``` is compatible with the list output from the ```combineTCR()``` function from the [scRepertoire](https://github.com/ncborcherding/scRepertoire) R package, while ```runTrex()``` must be performed on a single-cell object.

**chains**  
* "TRA" for TCR Alpha Chain
* "TRB" for TCR Beta Chain

**method**  
* "encoder" for a convolution neural network (CNN) based encoding. 
* "geometric" for a geometric transformation  

**encoder.model**  
* "VAE" for a variational autoencoder  
* "AE" for a traditional autoencoder  

**encoder.input**  
* "AF" to use Atchley factors  
* "KF" to use Kidera factors  
* "both" to use both  
* "OHE" for a One Hot Autoencoder

**theta**  
If choosing the geometric transformation, what value of theta to use (default is pi)

```{r}
library(Trex)
Trex_vectors <- maTrex(scRep_example, 
                       chains = "TRA",
                       encoder.model = "VAE", 
                       encoder.input = "AF")

ggplot(data = as.data.frame(Trex_vectors), aes(Trex_1, Trex_2)) + 
  geom_point() + 
  theme_classic()

Trex_vectors2 <- maTrex(scRep_example, 
                       chains = "TRB",
                       method = "geometric",
                       theta = pi)

ggplot(as.data.frame(Trex_vectors2), aes(x = Trex_1, y = Trex_2)) + 
  geom_point() + 
  theme_classic()
```

## runTrex

Additionally, ```runTrex()``` can be used to append the Seurat or Single-cell Experiment object with the Trex vectors and allow for further analysis. Importantly, ```runTrex()``` will remove single cells that do not have recovered TCR data in the metadata of the object. 

```{r}
scRep_example <- runTrex(scRep_example, 
                         chains = "TRB",
                         encoder.model = "VAE", 
                         encoder.input = "KF",
                         reduction.name = "Trex.KF")
```


# Using Trex Vectors

After ```runTrex()``` we have the encoded values stored under **"Trex..."**. Using the Trex reduction stored in Seurat, we can calculate the nearest neighbor and shared nearest neighbor indexes and generate a UMAP.

```{r}
#Generating UMAP from Trex Neighbors
scRep_example <- RunUMAP(scRep_example, 
                         reduction = "Trex.KF",
                         dims = 1:30,
                         reduction.name = 'Trex.umap', 
                         reduction.key = 'trexUMAP_')

#Trex UMAP
plot1 <- DimPlot(scRep_example, 
                 reduction = "Trex.umap") + 
         NoLegend()
plot2 <- DimPlot(scRep_example, 
                 group.by = "CTaa", 
                 reduction = "Trex.umap") + 
          scale_color_viridis(discrete = TRUE, option = "B") + 
          NoLegend()

plot1 + plot2
```

We now can use this in a similar way as other single-cell modalities and calculate weighted nearest neighbor (WNN). To check out more on WNN, please read the Satija's group [paper](https://pubmed.ncbi.nlm.nih.gov/34062119/). We will use the RNA and Trex vectors for the WNN calculations. 

```{r}
scRep_example <- FindMultiModalNeighbors(
                    scRep_example, 
                    reduction.list = list("pca", "Trex.KF"), 
                    dims.list = list(1:30, 1:30), 
                    modality.weight.name = "RNA.weight")

scRep_example <- RunUMAP(scRep_example, 
                     nn.name = "weighted.nn", 
                     reduction.name = "wnn.umap", 
                     reduction.key = "wnnUMAP_")

scRep_example <- FindClusters(scRep_example, 
                          graph.name = "wsnn", 
                          resolution = 0.6,
                          algorithm = 3, verbose = FALSE)

#WNN UMAP
plot3 <- DimPlot(scRep_example, 
                 reduction = "wnn.umap")
plot4 <- DimPlot(scRep_example, 
                 reduction = "wnn.umap", 
                 group.by = "CTaa") + 
          scale_color_viridis(discrete = TRUE, option = "B") + 
          NoLegend()

plot3 + plot4
```

## Comparing the outcome to just one modality

We can also look at the differences in the UMAP generated from RNA or Trex as individual components. Remember, the clusters that we are displaying in UMAP are based on clusters defined by the weighted nearest neighbors calculated above. 

```{r}
scRep_example <- RunUMAP(scRep_example, 
                         reduction = 'pca', 
                         dims = 1:30, 
                         assay = 'RNA', 
                         reduction.name = 'rna.umap', 
                         reduction.key = 'rnaUMAP_')


plot5 <- DimPlot(scRep_example, reduction = "rna.umap") + NoLegend()
plot6 <- DimPlot(scRep_example, reduction = "Trex.umap") + NoLegend()

plot5 + plot6
```


# CoNGA Reduction

Recent [work](https://pubmed.ncbi.nlm.nih.gov/34426704/) has proposed using representative cells for the characterization of clonotype and gene expression relationships. In order to generate these representative cells, either a mean expression across a clone or using the PCA dimensional space to identify a single cell that has the minimum euclidean distance across a clone. 

In order to generate a single-cell object based on the CoNGA approach, Trex offers the function ```CoNGAfy()```. For **method**, select either "mean" or "dist" as described above. After performing ```CoNGAfy()```, the user can use any of the above reduction strategies. 

```{r}
CoNGA.seurat <- CoNGAfy(scRep_example, 
                         method = "dist")

CoNGA.seurat <- runTrex(CoNGA.seurat, 
                        chains = "TRB",
                        encoder.model = "VAE", 
                        encoder.input = "KF",
                        reduction.name = "Trex.KF")

CoNGA.seurat <- CoNGA.seurat %>%
                  FindNeighbors(reduction = "Trex.KF") %>%
                  FindClusters(algorithm = 3)

CoNGA.seurat <- RunUMAP(CoNGA.seurat, 
                        reduction = "Trex.KF", 
                        dims = 1:20, 
                        reduction.name = 'Trex.umap', 
                        reduction.key = 'trexUMAP_')

DimPlot(CoNGA.seurat, reduction = "Trex.umap") + NoLegend()
```

# Annotating possible epitopes

Towards find epitope relationships, Trex has a built in data base of TCRA and TCRB sequences associated with epitopes. To append the database to the single-cell object (either before or after ```CoNGAfy()```), you can use ```annotateDB()```. 

The database is collated from multiple sources and represents a great deal of effort from these sources:

1. [VDJdb](https://vdjdb.cdr3.net/) - [citation](https://pubmed.ncbi.nlm.nih.gov/28977646/)
2. [McPAS-TCR](http://friedmanlab.weizmann.ac.il/McPAS-TCR/) - [citation](https://pubmed.ncbi.nlm.nih.gov/28481982/)
3. [iedb](https://www.iedb.org/) - [citation](https://pubmed.ncbi.nlm.nih.gov/30357391/)
4. [TBA](https://db.cngb.org/pird/home/) (now called PIRD) - [citation](https://pubmed.ncbi.nlm.nih.gov/31373607/)


```{r}
CoNGA.seurat <- annotateDB(CoNGA.seurat, 
                           chains = "TRB")

DimPlot(CoNGA.seurat, 
        reduction = "Trex.umap", 
        group.by = "TRB_Epitope.species")

DimPlot(CoNGA.seurat, 
        reduction = "Trex.umap", 
        group.by = "TRB_Epitope.sequence")
```

Or ```annotateDB()``` can be used on the full single-cell object and examine the sequence information along the RNA-based UMAP. An added feature to the function allows the annotations to be extended to cdr3 sequences that are within defined edit distances from the reference using **edit.distance**.

```{r}
scRep_example <- annotateDB(scRep_example,
                        chains = "TRB", 
                        edit.distance = 2)

DimPlot(scRep_example, 
        reduction = "rna.umap", 
        group.by = "TRB_Epitope.species")
```

